\chapter{General Introduction}

\label{chap:Intro}

This thesis presents work investigating the role that recombination plays in the adaptive evolution of two eukaryotic microorganisms, \textit{Albugo candida} and \textit{Fragilariopsis cylindrus}.
Both of these organisms exist in environments that may be considered very dynamic.

In addition, methodological work was also conducted which implemented and tested software dedicated to making it easier to detect recombination in Next Generation Sequencing data. 
The software was also designed to help solve current methodological issues with distinguishing mosaic regions that are the result of hybridisation, and those that are the result of incomplete lineage sorting.

These works are presented in chapters \ref{chap:HC}, \ref{chap:Acandida}, and \ref{chap:diatom}. Each has a more detailed and focused introduction to the concepts specific to them.
It is the purpose of this chapter to provide an overview of the key concepts of population genetics that are relevant to this work and provide a wider context for the next three chapters.

In order to understand adaptive evolution, it is necessary to understand the five forces of population genetics and how they drive adaptive evolution.
What follows is an overview of the five fundamental forces of evolutionary change. Afterwards, an overview of hybrid zones, and an overview of current common Bioinformatics procedures and how they are used in population genetics analyses are presented.   

\section{The five forces of evolutionary change}

\input{Chapters/Introduction/Selection.tex}
\input{Chapters/Introduction/Drift.tex}
\input{Chapters/Introduction/Mutation.tex}
\input{Chapters/Introduction/Gene_Flow.tex}
\input{Chapters/Introduction/Recombination.tex}
\input{Chapters/Introduction/HybridZones.tex}


\section{The role of Bioinformatics in population genetics}

\input{Chapters/Introduction/Bioinformatics.tex}
